---
title: "DEG_4, p=0.1"
format: html
editor: visual
---

# SETUP

```{r, setup, include=FALSE}

filespath = "C:/Users/Simo/genomics/damianoRNAseq-20042023/output/assembly_genomeref" # set your own path
#setwd(filespath)
knitr::opts_knit$set(root.dir = filespath)

```

## Libraries

```{r,libraries}
#| include: false
#|output: false
library(tidyverse)
library(dplyr)
library(ggrepel) #labels
library(ggbreak)
library(ggpubr)

library(edgeR) # DE
library(DescTools) # sql operator %like% for easy peasy qwery
library(GeneOverlap)

# magic plots
library(viridis)
library(ComplexHeatmap)
library(circlize)

library(patchwork)

library(gprofiler2)



getwd()
```

## Data load and preparation

```{r, pre-preprocessing}

group <- factor(c("WT", "WT",
                  "eel", "eel", "eel",
                  "areb3_1", "areb3_1", "areb3_1",
                  "eel.areb3_1", "eel.areb3_1", "eel.areb3_1"), 
           levels = c("WT", "eel", "areb3_1", "eel.areb3_1"))
#group <- relevel(group, ref = "WT")


labels <- c("A3","A4",
            "B1","B2","B3",
            "C1","C2","C4",
            "D1","D2","D3")

lane <- factor(c(4,4,
                 1,1,1,
                 4,4,1,
                 1,1,1))

count_files <- grep("_unstranded.txt", list.files(), value = T)

count_files <-  count_files[-1]

y <- readDGE(count_files, group = group)
y$samples$lane <- lane
y$samples$group <- relevel(y$samples$group, ref = "WT")


```

```{r}
clean_sample_names <- function(count_table, split_arg) { # takes a named vector and renames it
  lapply(colnames(count_table) %>%    
           strsplit(colnames(count_table), split = split_arg),
                        FUN = '[', 1)
}

colnames(y) <- clean_sample_names(y, '_EK')
#colnames(y)

# palette for plots 
group_colors <- c(rep("#24d4a3",2),
                  rep("#e84d8a",3),
                  rep("#64c5eb",3),
                  rep("#7f58af",3))

```

### Retrive gene names for conversion

```{r}

#gene_ids_unfiltered <- rownames(y)
#gene_ids
#write.table(gene_ids_unfiltered, "genes_ids_unfiltered.txt", row.names = F, col.names = F)
#write.csv(y$counts, "Damiano_RAW.csv")

```

## Add some layes of metadata

### sequencing quality data

```{r}

# read metadata file
meta_DE <- read_csv("metadata/DE_metadata.csv", show_col_types = F )
#colnames(meta_DE)
meta_DE <- meta_DE[-1,]

# add metadata to the DGE object
# actually looks kinda useless BUT it may work for smth
y$samples$n_input_reads <- unlist(meta_DE[,2])
y$samples$avg_read_len <- unlist(meta_DE[,3]) # ma è la stessa per tutti
y$samples$uniquely_mapped_reads_n <- unlist(meta_DE[,4]) 
y$samples$uniquely_mapped_reads_pcent <- unlist(meta_DE[,5])

y$samples$n_splices_total <- unlist(meta_DE[,6])
y$samples$n_splices_ann_sjdb <- unlist(meta_DE[,7])
y$samples$n_splices_noncanon <- unlist(meta_DE[,8])

y$samples$n_reads_mapped_multi_loci <- unlist(meta_DE[,9])
y$samples$pcent_reads_mapped_multi_loci <- unlist(meta_DE[,10])

y$samples$n_reads_mapped_toomany_loci <- unlist(meta_DE[,11])
y$samples$pcent_reads_mapped_toomany_loci <- unlist(meta_DE[,12])

y$samples$n_reads_unmapped_toomany_mm <- unlist(meta_DE[,13])
y$samples$pcent_reads_unmapped_toomany_mm <- unlist(meta_DE[,14])

y$samples$n_reads_unmapped_too_short <- unlist(meta_DE[,15])
y$samples$pcent_reads_unmapped_too_short <- unlist(meta_DE[,16])

y$samples$n_reads_unmapped_other <- unlist(meta_DE[,17])
y$samples$pcent_reads_unmapped_other <- unlist(meta_DE[,18])


```

### Genes metadata

```{r}
#| include: false
#|output: false

# metadata/ varie liste di conversion

entrez_convers_file1 <- read_table("metadata/TAIR_to_NCBI.txt", 
                                   col_names = F)

entrez_convers_file2 <- read_table("metadata/entrez_conv.txt", 
                                   col_names = T)

entrez_convers_file2  <- entrez_convers_file2[,c(1,2,5)]



entrez_convers_file3 <- read_table("metadata/officialGeneName_conv.txt",
                                   col_names = T)

entrez_convers_file3[, 1:2] -> entrez_convers_file3



TAIR_IDs <- data.frame(rownames(y))  
colnames(TAIR_IDs) <- "TAIR_ID"




left_join(data.frame(TAIR_ID = TAIR_IDs), entrez_convers_file1, by = join_by(TAIR_ID == X2)) %>%
  rename(.,c(Entrez_ID = X1))-> genes_meta_1


left_join(data.frame(TAIR_ID = TAIR_IDs), entrez_convers_file2, by = join_by(TAIR_ID == From))%>% 
  rename(.,c(Entrez_ID = To, CommonName = Name)) -> genes_meta_2

left_join(data.frame(TAIR_ID = TAIR_IDs), entrez_convers_file3, by = join_by(TAIR_ID == From))%>% 
  rename(.,c(Name = To)) -> genes_meta_3



genes_meta <- left_join(genes_meta_1, genes_meta_3) %>%
              left_join(., genes_meta_2)




#table(is.na(genes_meta_before[2]))# 5491 NAs # we have some na to inspect after filtering

y$genes <- genes_meta

# i need TAIR | Entrez | OtherName | common
# write.csv(genes_meta, "metadata/Converted_gene_Names.csv")

```

## 

# Design matrix

```{r, design_matrix}

design <- model.matrix(~0+group, data = y$samples$group)
colnames(design) <- levels(y$samples$group)

```

# Filtering

\*From EdgeR manual:

*The function keeps rows that have worthwhile counts in a minimum number of samples. It accesses the group factor contained in y in order to compute the minimum group size, but the filtering is performed independently of which sample belongs to which group so that no bias is introduced. It is recommended to recalculate the library sizes of the DGEList object after the filtering, although the downstream analysis is robust to whether this is done or not*

filtering function has actually 2 filters, one based on raw counts and one on cpm. To be filtered, a feature has to pass both filters \[*source: i read the source code*\]

```{r}

dim(y) # 33609
```

```{r}

table(filterByExpr(y, design))
table(rowSums(y$counts>=5)>= 2) 

keep <- rowSums(y$counts>=5)>= 2

y2 <- y[keep,, keep.lib.sizes=F]
dim(y2) # 21238
```

### Genes lost after filtering

```{r}

dim_after_filter <- dim(y2)
dim_before_filter <- dim(y)

dim_after_filter
dim_before_filter

dim_before_filter[1]-dim_after_filter[1] # we loose  genes with filtering


```

### 

### Table of genes after filtering -\> conversion

```{r}

#gene_ids_filtered <- rownames(y)
#gene_ids
#write.table(gene_ids_filtered, 
 #           "genes_ids_filtered.txt", 
  #          row.names = F, 
   #         col.names = F)
```

### genes meta after filter

```{r}

# metadata/ varie liste di conversion

#entrez_convers_file <- read_table("metadata/TAIR_to_NCBI.txt", col_names = F)


#TAIR_IDs <- data.frame(rownames(y2)) %>% rename(.,c(TAIR_ID = rownames.y2.))


#data.frame(TAIR_ID = entrez_convers_file[2], Entrez_ID = entrez_convers_file[1]) %>% rename(c(TAIR_ID = X2, Entrez_ID = X1)) %>%
 # left_join(data.frame(TAIR_ID = TAIR_IDs), .) -> genes_meta_after

#table(is.na(genes_meta_after[2]))# 5491 NAs # we have some na to inspect after filtering

## add common names

#common_names_file <- read.csv("metadata/")
```

## Recalc of lib size

Recomputing the library size after filtering is not a standard practice in edgeR *or RNA-Seq analysis in general.*

Once the library sizes are initially computed and normalized, they are typically considered fixed for the analysis. Recomputing library sizes could *potentially introduce bias or complications in downstream analyses*.

# Normalization

### `calcNormFactors` normalizes the library sizes.

### `cpm` divides counts by library sizes.

*LogCPM is calculated dividing the number of reads of a gene for the total of the reads in the sample, then multiply for a million and then apply a log2 transformation. This normalize the data about expression based on the dimension of the sample, allowing for a more accurate comparison between samples of different sizes.*

[\[bioc forum\]](https://support.bioconductor.org/p/9145189/#9145192) Running `cpm` after `calcNormFactors` uses normalized library sizes. Running `cpm` before `calcNormFactors` uses unnormalized library sizes.

Obviously the **former is better than the latter**. Why would you normalize the library sizes but then ignore the normalization? That would just make no sense at all.

```{r}

logcpm_before <- cpm(y2, log=TRUE) #log before norm
logmed_b <- apply(logcpm_before,2, FUN = median)

y2 <- calcNormFactors(y2, method = "TMM") ###### Normalization ###########


logcpm_after <- cpm(y2, log=TRUE) #  after norm
logmed_a <- apply(logcpm_after,2, FUN = median)

logmed_mat <- matrix(c(logmed_b, logmed_a), nrow = 2, byrow = T)
colnames(logmed_mat) <- group
rownames(logmed_mat) <- c('Before', 'After')
t(logmed_mat)

par(mfrow = c(1,2))
boxplot(logcpm_before, 
        ylab = 'log(CPM)', 
        xlab = 'samples', 
        col = group_colors, 
        alpha = .75,
        main= 'Before')


boxplot(logcpm_after, 
        xlab = 'samples', 
        col = group_colors, 
        main = 'After')
title("TMM Normalization", line = -1, outer = TRUE)

```

```{r}

#write.csv(cpm(y2$counts), 'CPM_Damiano_noA1.csv')

```

## Vizualization of normalization

```{r}
data.frame(Sample = colnames(y2), 
           Norm_Factor = y2$samples$norm.factors-1) %>%
  ggplot(., aes(x = Sample, y = Norm_Factor)) +
  geom_bar(stat = "identity", fill = group_colors, alpha = .75) +
    #coord_cartesian(ylim = c(min(y2$samples$norm.factors), max(y2$samples$norm.factors)))+

  # Add labels and a title
  labs(
    x = "Sample",
    y = "Scaling",
    title = "Visual representation of scaling factors"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1))


```

```{r}

data.frame(Normalization_Factor = y2$samples$norm.factors,
           row.names = colnames(y))

# plot
data.frame(Normalization_Factor = y2$samples$norm.factors,
           row.names = colnames(y2)) %>%
  ggplot(data = ., aes(x = rownames(.), y = Normalization_Factor)) + 
  geom_bar(colour = "black",
           fill = group_colors,
           alpha = .75,
           stat = 'identity') + theme_minimal()
```

MultiDimensional Scaling plot

```{r}
plotMDS(y2, 
        labels = group, 
        col = group_colors,
        main = "MDS - expression profile distance among genotypes")

plotMDS(y2, 
        labels = colnames(y2), 
        col = group_colors,
        main = "MDS - expression profile distance among genotypes")
```

# 

# Dispesion and BCV

```{r}

y2 <- estimateDisp(y2,design, robust = F) 
plotBCV(y2)

```

# Compare groups:

### Classic approach: Exact Test - less powerful

### \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

## GLM & Quasi-likelyhood F-test

When you specify a contrast "A vs B", the contrast coefficients are set as

### **-1 \* condition_A + 1 \* condition_B**

Positive logFC indicates that the gene is upregulated in condition B compared to condition A. Conversely, a negative logFC indicates upregulation in condition A compared to condition B

(or downregulated in condition B relative to condition A)

## 

### Constrasts

```{r}
# first define contrasts

my_constrasts <- makeContrasts(WT_eel = eel-WT,
                              WT_areb3 = areb3_1-WT,
                              WT_double = eel.areb3_1-WT,
                              areb3_eel= eel-areb3_1,
                              areb3_double = eel.areb3_1-areb3_1,
                              eel_double = eel.areb3_1-eel,
                              levels = design) # adjust for intercept

data.frame(my_constrasts)
```

# Model fitting

-   glm for negative binomial

```{r}
fit_nointercept <- glmQLFit(y2,design, robust = T, legacy = F)
plotQLDisp(fit_nointercept)
```

# Pairwise comparisons

```{r}

# WT vs areb3_1
qlf_wtareb <- glmQLFTest(fit_nointercept, 
                         contrast=my_constrasts[,"WT_areb3"])
results_wtareb_unfiltered <- topTags(qlf_wtareb,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_wtareb, p.value=0.05))
data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_wtareb, p.value=0.05))[1],
                     summary(decideTests(qlf_wtareb, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "areb3_1 vs WT"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_wtareb



# wt vs eel
qlf_wteel <- glmQLFTest(fit_nointercept, 
                        contrast=my_constrasts[,"WT_eel"])
results_wteel_unfiltered <- topTags(qlf_wteel,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_wteel, p.value=0.05))
data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_wteel, p.value=0.05))[1],
                     summary(decideTests(qlf_wteel, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "eel vs WT"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_wteel





qlf_wtdouble <-glmQLFTest(fit_nointercept,
                          contrast=my_constrasts[,"WT_double"])
results_wtdouble_unfiltered <- topTags(qlf_wtdouble,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_wtdouble, p.value=0.05))

data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_wtdouble, p.value=0.05))[1],
                     summary(decideTests(qlf_wtdouble, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "Double mutant vs WT"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_wtdouble






qlf_arebeel <- glmQLFTest(fit_nointercept,
                          contrast=my_constrasts[,"areb3_eel"])
results_arebeel_unfiltered <- topTags(qlf_arebeel,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_arebeel, p.value=0.05))
data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_arebeel, p.value=0.05))[1],
                     summary(decideTests(qlf_arebeel, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "eel vs areb3_1"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_arebeel





qlf_arebdouble <- glmQLFTest(fit_nointercept, 
                             contrast=my_constrasts[,"areb3_double"])
results_arebdouble_unfiltered <- topTags(qlf_arebdouble,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_arebdouble, p.value=0.05))
data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_arebdouble, p.value=0.05))[1],
                     summary(decideTests(qlf_arebdouble, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "Double mutant vs areb3_1"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_arebdouble





qlf_eeldouble <- glmQLFTest(fit_nointercept, 
                            contrast=my_constrasts[,"eel_double"])
results_eeldouble_unfiltered <- topTags(qlf_eeldouble,
        n = Inf,
        adjust.method = "BH", 
        sort.by = "PValue", 
        p.value = 1)
summary(decideTests(qlf_eeldouble, p.value=0.05))
data.frame(Regulation = c("Down", "Up"), 
           Genes = c(summary(decideTests(qlf_eeldouble, p.value=0.05))[1],
                     summary(decideTests(qlf_eeldouble, p.value=0.05))[3])) %>%
  ggplot(., aes(x = Regulation, y = Genes)) +
  geom_bar(stat = "identity", fill = c("#114155", "#3b97b6")) +
  labs(
    x = "Regulation",
    title = "Significant DE genes",
    subtitle = "Double mutant vs eel"
  )+
  theme_minimal()+
    theme(axis.text.x = element_text(angle = 0, 
                                   hjust = .8, 
                                   vjust = 1)) -> plot_eeldouble



# test all  6 contrasts combinations

#plot_wtareb+plot_wteel+plot_wtdouble
#plot_arebeel+plot_arebdouble+plot_eeldouble

#plot_wtareb+plot_wteel+plot_wtdouble+ plot_arebeel+plot_arebdouble+plot_eeldouble
plot_wtareb
plot_wteel
plot_wtdouble
plot_arebeel
plot_arebdouble
plot_eeldouble
```

# 

# Function that filters the topTag genes

from the results of topTags, this function selects genes that have significant and impactant DE:

-   p.value (adjusted) \< .1

-   significant expression

```{r}

extract_genes <- function(topTags_result,FDR_ = 0.05, logfc_ = 1  ) {
  
  topTags_result <- data.frame(topTags_result)
  
  topTags_result <- topTags_result[which((topTags_result$logFC > logfc_|     
                                          topTags_result$logFC < -logfc_ ) & 
                                          topTags_result$FDR < FDR_ & 
                                          topTags_result$logCPM > 0),]
  
  return(topTags_result)

  
}

results_wtareb <- extract_genes(results_wtareb_unfiltered)

results_wteel <- extract_genes(results_wteel_unfiltered)

results_wtdouble <- extract_genes(results_wtdouble_unfiltered)

results_arebeel <-  extract_genes(results_arebeel_unfiltered)

results_arebdouble <- extract_genes(results_arebdouble_unfiltered)

results_eeldouble <- extract_genes(results_eeldouble_unfiltered)

# _unfiltered for volcanoes
```

by intersecting results from previus cell we can get list of most varying genes for each genotype/group

```{r}
# intersection
top_wt <- intersect(rownames(results_wtareb), c(rownames(results_wteel), rownames(results_wtdouble)))
top_areb <- intersect(rownames(results_wtareb), c(rownames(results_arebeel), rownames(results_arebdouble)))
top_eel <- intersect(rownames(results_wteel), c(rownames(results_arebeel), rownames(results_eeldouble)))
top_double <- intersect(rownames(results_wtdouble), c(rownames(results_eeldouble), rownames(results_arebdouble)))

```

### Marker genes for wild type

```{r}
# expression of a gene across samples
markers_wt <- t(cpm(y2[which(rownames(y2) %in% top_wt),],log = T)) %>% data.frame(.)
```

```{r}
#library(ComplexHeatmap)
Heatmap(t(markers_wt))
```

### Marker genes for areb

```{r}

markers_areb <- t(cpm(y2[which(rownames(y2) %in% top_areb),],log = T)) %>% data.frame(.)
Heatmap(t(markers_areb))

```

### Marker genes for eel

```{r}

markers_eel <- t(cpm(y2[which(rownames(y2) %in% top_eel),],log = T)) %>% data.frame(.)
Heatmap(t(markers_eel))

```

### Marker genes for double mutant

```{r}

markers_double <- t(cpm(y2[which(rownames(y2) %in% top_double),],log = T)) %>% data.frame(.)
Heatmap(t(markers_double))

# we have only 1 gene even being poorly restrictive
```

### 

# MD ggplot2

### Plot function

```{r}
###########################################à

plot_MD <- function(Toptags_results, FDR_ = .05, logFC_ = 1){
  Toptags_results <- Toptags_results$table
  
  regulation <- ifelse(Toptags_results$FDR < FDR_ & Toptags_results$logFC > logFC_, "UP", 
                ifelse(Toptags_results$FDR < FDR_ & Toptags_results$logFC < -logFC_, "DOWN", 
                "none"))
  
  labelled_genes <-  subset(Toptags_results, abs(Toptags_results$logFC) > logFC_ & Toptags_results$FDR < FDR_)

  data.frame(Toptags_results) %>%
  ggplot(., 
       aes(x = logCPM, 
           y=logFC,
           col=regulation,
           size = ifelse(regulation %in% c("UP", "DOWN"), 2, 1)), 
           alpha=.75) + 
  geom_point() + 
  scale_colour_manual(values=c("blue", "black","red"))+
    #scale_size_continuous("FDR") +
     #   scale_size_continuous("logFC")+

  geom_text_repel(data = labelled_genes,
            aes(label = rownames(labelled_genes)),
            col = "black", 
            size = 3.5, 
            nudge_x = 0, 
            nudge_y = ifelse(regulation == "UP", 2, 
                             ifelse(regulation == "DOWN", -2, 0)), 
            #max.overlaps = Inf,
            force = 1) +
    theme_minimal()+
    
  guides(size = "none")
  
  
}
```

## MD for mutants vs WT

```{r}

plot_MD(results_wtareb_unfiltered) + 
  labs(title = "Mean-difference plot", subtitle = "areb vs WT")

plot_MD(results_wteel_unfiltered) +
  labs(title = "Mean-difference plot", subtitle = "eel vs WT")

plot_MD(results_wtdouble_unfiltered) +
  labs(title = "Mean-difference plot", subtitle = "Double mutant vs WT")



```

## MD for mutants vs mutants

```{r}

plot_MD(results_arebeel_unfiltered,FDR_ = 0.1, logFC_ = .5)+ 
  labs(title = "Mean-difference plot", subtitle = "eel vs areb3")

plot_MD(results_arebdouble_unfiltered,FDR_ = 0.1, logFC_ = .5)+ 
  labs(title = "Mean-difference plot", subtitle = "Double mutant vs areb3")

plot_MD(results_eeldouble_unfiltered,FDR_ = 0.1, logFC_ = .5) + 
  labs(title = "Mean-difference plot", subtitle = "Double mutant vs eel")


```

# Contrasts

#### Wild Type vs areb3

```{r}

results_wtareb

t(cpm(y2[which(rownames(y2) %in% rownames(results_wtareb)),c(1:2,6:8)],
      log = T)) %>% t() %>% data.frame(.) %>%
Heatmap(.)
```

#### Wild Type vs eel

```{r}
results_wteel

t(cpm(y2[which(rownames(y2) %in% rownames(results_wteel)),c(1:5)],
      log = T)) %>% t() %>% data.frame(.) %>%
Heatmap(.)
```

#### Wild type vs double mutant

```{r}

results_wtdouble

t(cpm(y2[which(rownames(y2) %in% rownames(results_wtdouble)),c(1:2,9:11)],
      log = T)) %>% t() %>% data.frame(.) %>%
Heatmap(.)
```

```{r}

rownames(results_wtareb) %>% data.frame()
rownames(results_wteel) %>% data.frame()
rownames(results_wtdouble) %>% data.frame()


results_wtareb[order(abs(results_wtareb$logFC), decreasing = T),]
results_wteel[order(abs(results_wteel$logFC), decreasing = T),]
results_wtdouble[order(abs(results_wtdouble$logFC), decreasing = T),]


```

overlap with tfs

```{r}

read_table('Ath_TF_list.txt') -> tf_list

table(unlist(tf_list['Gene_ID']) %in% rownames(y2$counts)
) 
# so we count 1886 tfs out of 2296


```

```{r}
# differentially expressed TFs
results_wtareb[which(rownames(results_wtareb) %in% 
                       unlist(tf_list['Gene_ID'])),] -> areb3_tfs
areb3_tfs

results_wteel[which(rownames(results_wteel) %in% 
                       unlist(tf_list['Gene_ID'])),]

results_wtdouble[which(rownames(results_wtdouble) %in% 
                       unlist(tf_list['Gene_ID'])),]
```

```{r}
gost(areb3_tfs$TAIR_ID,
     organism = 'athaliana',
     multi_query = F, 
     evcodes = T,
     highlight = T,
     exclude_iea = T,
     custom_bg = background_genes) -> areb3_tfs_enrich

plot_from_gost(areb3_tfs_enrich, annDB = 'GO:BP', top_res = 20, highlighted_only = T) 
plot_from_gost(areb3_tfs_enrich, annDB = 'GO:BP', top_res = 20)

plot_from_gost(areb3_tfs_enrich, annDB = 'GO:MF', top_res = 20, highlighted_only = T) 
plot_from_gost(areb3_tfs_enrich, annDB = 'GO:MF', top_res = 20)  
#ggsave('mfplotD.png', path = filespath )

plot_from_gost(areb3_tfs_enrich, annDB = 'GO:CC', top_res = 20, highlighted_only = T) 
#plot_from_gost(areb3_tfs_enrich, annDB = 'GO:CC', top_res = 20)

plot_from_gost(areb3_tfs_enrich, annDB = 'WP', top_res = 20) 
#plot_from_gost(areb3_tfs_enrich, annDB = 'KEGG', top_res = 20) 
```

```{r}
# group A bzips expression profile

a_bZips <- list("AT2G17770","AT4G35900","AT1G03970",
                "AT5G44080","AT2G41070","AT3G56850",
                "AT2G36270","AT3G44460","AT5G42910",
                "AT1G45249","AT4G34000","AT1G49720","AT3G19290")


s_bZips <- list("AT5G60830","AT5G08141","AT1G68880","AT3G30530",
                "AT5G38800","AT5G15830","AT1G13600","AT2G04038",
                "AT3G62420","AT2G18160","AT1G75390","AT4G34590",
                "AT5G49450","AT1G59530","AT3G49760","AT2G22850","AT4G37730")

c_bZips <- list("AT3G54620","AT4G02640","AT5G28770","AT5G24800")

d_bZips <- list("AT1G22070","AT1G77920","AT5G65210","AT5G10030",
                "AT1G08320","AT5G06839","AT1G68640","AT5G06960",
                "AT5G06950","AT3G12250")

f_bZips <- list("AT4G35040", "AT2G16770","AT3G51960")

i_bZips <- list("AT1G06850","AT2G40620","AT1G43700","AT1G06070",
                "AT2G31370","AT2G21230","AT4G38900","AT2G24340",
                "AT2G21235","AT2G12900","AT2G13150","AT2G12940")


e_bZips <- list("AT2G42380","AT3G58120","AT5G04840",
                "AT1G35490","AT1G58110","AT4G06598")


b_bZips <- list("AT3G56660","AT2G40950","AT3G10800")


jhkm_bZips <- list("AT1G42990","AT5G11260","AT3G17609","AT1G19490","AT5G07160")

g_bZips <- list("AT2G46270","AT4G01120","AT4G36730","AT1G32150","AT2G35530")

```

```{r}
which(a_bZips %in% rownames(y2)) # missing 8 and 9 = AT3G44460, AT5G42910
cpm(y2$counts[which(rownames(y2) == "AT3G56850"), ])



cpm(y2$counts[which(rownames(y2) %in% a_bZips), ]) %>% data.frame() -> a_bZips_expression

results_wtareb[which(rownames(results_wtareb) %in% 
                       rownames(a_bZips_expression)),]

results_wteel[which(rownames(results_wteel) %in% 
                       rownames(a_bZips_expression)),]

results_wtdouble[which(rownames(results_wtdouble) %in% 
                       rownames(a_bZips_expression)),]

```

```{r}
#plot per genotye

#colnames(a_bZips_expression) <- as.factor(colnames(a_bZips_expression))
Heatmap(a_bZips_expression, column_order = sort(colnames(a_bZips_expression)))
```

```{r}

stack(a_bZips_expression) %>% 
  data.frame(Gene = rep(rownames(a_bZips_expression), length(a_bZips_expression[1,])),
             Genotype = sort(rep(group, 11),decreasing = F)) %>%
  rename(CPM = values, Sample = ind) -> stacked_abzip_express



ggplot(stacked_abzip_express, aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_wrap(~Gene)


```

```{r}

stacked_abzip_express[stacked_abzip_express$Gene %in% 
                        c("AT1G49720","AT1G45249"),]%>% 
  ggplot(., aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_wrap(~Gene)



stacked_abzip_express[stacked_abzip_express$Gene %in% 
                        c("AT4G34000", "AT3G19290"),]%>% 
  ggplot(., aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_wrap(~Gene)




stacked_abzip_express[stacked_abzip_express$Gene %in% 
                        c("AT1G49720","AT1G45249","AT4G34000", "AT3G19290"),]%>% 
  ggplot(., aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_grid(~Gene)




# fd, fdp, 
stacked_abzip_express[stacked_abzip_express$Gene %in% 
                        c("AT4G35900", "AT2G17770",),]%>% 
  ggplot(., aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_wrap(~Gene)

```

```{r}

# let's make some heatmaps for every group 


# A
cpm(y2$counts[which(rownames(y2) %in% a_bZips), ]) %>% data.frame() -> expr_A
y2[y2$genes$TAIR_ID %in% rownames(expr_A),]$genes$Name -> names_A
rownames(expr_A) <- names_A
expr_A %>% as.matrix() %>% Heatmap(., column_title = 'Group A expression')

# B
cpm(y2$counts[which(rownames(y2) %in% b_bZips), ]) %>% data.frame() -> expr_B
y2[y2$genes$TAIR_ID %in% rownames(expr_B),]$genes$Name -> names_B
rownames(expr_B) <- names_B
expr_B %>% as.matrix() %>% Heatmap(., column_title = 'Group B expression')


# C 
cpm(y2$counts[which(rownames(y2) %in% c_bZips), ]) %>% data.frame() -> expr_C
y2[y2$genes$TAIR_ID %in% rownames(expr_C),]$genes$Name -> names_C
rownames(expr_C) <- names_C
expr_C %>% as.matrix() %>% Heatmap(., column_title = 'Group C expression')


# D 
cpm(y2$counts[which(rownames(y2) %in% d_bZips), ]) %>% data.frame() -> expr_D
y2[y2$genes$TAIR_ID %in% rownames(expr_D),]$genes$Name -> names_D
rownames(expr_D) <- names_D
expr_D %>% as.matrix() %>% Heatmap(., column_title = 'Group D expression')


# E 
cpm(y2$counts[which(rownames(y2) %in% e_bZips), ]) %>% data.frame() -> expr_E
y2[y2$genes$TAIR_ID %in% rownames(expr_E),]$genes$Name -> names_E
rownames(expr_E) <- names_E
expr_E %>% as.matrix() %>% Heatmap(., column_title = 'Group E expression')


# F
cpm(y2$counts[which(rownames(y2) %in% f_bZips), ]) %>% data.frame() -> expr_F
y2[y2$genes$TAIR_ID %in% rownames(expr_F),]$genes$Name -> names_F
rownames(expr_F) <- names_F
expr_F %>% as.matrix() %>% Heatmap(., column_title = 'Group F expression')


# G
cpm(y2$counts[which(rownames(y2) %in% g_bZips), ]) %>% data.frame() -> expr_G
y2[y2$genes$TAIR_ID %in% rownames(expr_G),]$genes$Name -> names_G
rownames(expr_G) <- names_G
expr_G %>% as.matrix() %>% Heatmap(., column_title = 'Group G expression')


# I
cpm(y2$counts[which(rownames(y2) %in% i_bZips), ]) %>% data.frame() -> expr_I
y2[y2$genes$TAIR_ID %in% rownames(expr_I),]$genes$Name -> names_I
rownames(expr_I) <- names_I
expr_I %>% as.matrix() %>% Heatmap(., column_title = 'Group I expression')


# JHKM
cpm(y2$counts[which(rownames(y2) %in% jhkm_bZips), ]) %>% data.frame() -> expr_JHKM
y2[y2$genes$TAIR_ID %in% rownames(expr_JHKM),]$genes$Name -> names_JHKM
rownames(expr_JHKM) <- names_JHKM
expr_JHKM %>% as.matrix() %>% Heatmap(., column_title = 'Groups J/H/K/M expression')


# S
cpm(y2$counts[which(rownames(y2) %in% s_bZips), ]) %>% data.frame() -> expr_S
y2[y2$genes$TAIR_ID %in% rownames(expr_S),]$genes$Name -> names_S
rownames(expr_S) <- names_S
expr_S %>% as.matrix() %>% Heatmap(., column_title = 'Group S expression')








```

```{r}


```

```{r}

setwd("C:/Users/Simo/genomics/damianoRNAseq-20042023/meta_output_17d/meta_assembly_genomeref")
read.csv('FINAL_COUNT_TABLE_MAIDA.csv') -> y_maida
rownames(y_maida) <-  y_maida$X
y_maida <- y_maida[,-1]
maidagroup <- factor(c("WT", "WT", "WT",
                "fd", "fd", "fd",
                "fdp","fdp","fdp")) %>% relevel(., ref = "WT")

y_maida[which(rownames(y_maida) %in% a_bZips), ] %>% data.frame() -> a_bZips_expression_m

stack(a_bZips_expression_m) %>% 
  data.frame(Gene = rep(rownames(a_bZips_expression_m), length(a_bZips_expression_m[1,])),
             Genotype = sort(rep(maidagroup, 11),decreasing = F)) %>%
  rename(CPM = values, Sample = ind) -> stacked_abzip_express_m

ggplot(stacked_abzip_express_m, aes(x = Genotype, y = CPM, fill = Genotype))+
  geom_boxplot()+
  scale_fill_viridis_d()+
  theme_minimal() + facet_wrap(~Gene)

```

```{r}
# flowering time genes
library(readxl)

arabidopsis_tfs <- read_tsv("../Ath_TF_list.txt")
arabidopsis_tfs$Family %>% unique()

flowering_time_genes <- read_xls("../flowering_time_genes.xls", 
         sheet = 1, skip = 4)
flowering_time_genes$`Flowering pathway` %>% unique


getwd()
```

# ENRICHMENT

```{r}
# enrichment
plot_from_gost <- function(gost_result, 
                           annDB = c("GO:BP","GO:CC","GO:MF","KEGG","WP"),
                           #curated_only = F,
                           highlighted_only = F,
                           top_res = 10) {
  
  gost_result <- gost_result$result[gost_result$result$significant == T,]
  gost_result <- gost_result[gost_result$source %in% annDB,]
  
  if (highlighted_only == T) {
    gost_result <- gost_result[gost_result$highlighted == TRUE,]
  }
  
  gost_result <- gost_result %>% 
    group_by(query) %>%
    arrange(desc(intersection_size)) %>% 
    slice_head(n = top_res) %>% 
    ungroup()

gost_result$term_name <- factor(gost_result$term_name,
                                levels = unique(gost_result$term_name))

ggplot(gost_result, aes(x = intersection_size, 
                           y = term_name, 
                           fill = -log10(p_value) ))+
  scale_fill_viridis(option = 'mako', 
                     direction = -1,
                     #limits = c(0,50)
                     )+
  geom_bar(stat = 'identity') + 
  theme_minimal()+
  ggplot2::facet_grid(~query)+ labs(title = annDB)
}

```

```{r}


marker_results <- list(WT_Markers = top_wt,
                       areb3_Markers = top_areb,
                       eel_Markers = top_eel,
                       dbm_Markers = top_double)

de_results <- list(de_wtareb = rownames(results_wtareb),
                   de_wteel = rownames(results_wteel),
                   de_wtdbl = rownames(results_wtdouble))
background_genes <- rownames(y2$counts)
#data.frame(background_genes) %>% 
 # write_csv(., 'damianobg_p01.csv')

#data.frame(de_results[1])
#data.frame(de_results[2])
#data.frame(de_results[3])



```

```{r}


# enrich degs with custm bg
gost(de_results,
     organism = 'athaliana',
     multi_query = F, 
     evcodes = T,
     highlight = T,
     exclude_iea = T,
     custom_bg = background_genes) -> de_enriched_1

```

```{r}
de_enriched_1$result[de_enriched_1$result$query == 'de_wteel',]
de_enriched_1$result[de_enriched_1$result$query == 'de_wtdbl',]
de_enriched_1$result[de_enriched_1$result$query == 'de_wtareb',]
```

```{r}

plot_from_gost(de_enriched_1, annDB = 'GO:BP', top_res = 20, highlighted_only = T) 
plot_from_gost(de_enriched_1, annDB = 'GO:BP', top_res = 20)

plot_from_gost(de_enriched_1, annDB = 'GO:MF', top_res = 20, highlighted_only = T) 
plot_from_gost(de_enriched_1, annDB = 'GO:MF', top_res = 20)  
#ggsave('mfplotD.png', path = filespath )

plot_from_gost(de_enriched_1, annDB = 'GO:CC', top_res = 20, highlighted_only = T) 
plot_from_gost(de_enriched_1, annDB = 'GO:CC', top_res = 20)

plot_from_gost(de_enriched_1, annDB = 'WP', top_res = 20) 
plot_from_gost(de_enriched_1, annDB = 'KEGG', top_res = 20) 

```

```{r}
# DEGs with no BG

gost(de_results,
     organism = 'athaliana',
     multi_query = F, 
     evcodes = T,
     highlight = T,
     exclude_iea = T) -> de_enriched_nobg

```

```{r}


plot_from_gost(de_enriched_nobg, annDB = 'GO:BP', top_res = 20, highlighted_only = T) 
plot_from_gost(de_enriched_nobg, annDB = 'GO:BP', top_res = 30)

plot_from_gost(de_enriched_1, annDB = 'GO:MF', top_res = 20, 
               highlighted_only = T)

plot_from_gost(de_enriched_nobg, annDB = 'GO:MF', top_res = 20)  
#ggsave('mfplotD.png', path = filespath )

plot_from_gost(de_enriched_1, annDB = 'GO:CC', top_res = 20, 
               highlighted_only = T) 

plot_from_gost(de_enriched_nobg, annDB = 'GO:CC', top_res = 20)

plot_from_gost(de_enriched_nobg, annDB = 'WP', top_res = 20) 
plot_from_gost(de_enriched_nobg, annDB = 'KEGG', top_res = 20) 

```

```{r}
# markers 

gost(marker_results,
     organism = 'athaliana',
     multi_query = F, 
     evcodes = T,
     highlight = T,
     exclude_iea = T,
     custom_bg = background_genes) -> markers_enriched
```

```{r}


plot_from_gost(markers_enriched, annDB = 'GO:BP', top_res = 20, highlighted_only = T) 
plot_from_gost(markers_enriched, annDB = 'GO:BP', top_res = 20)

plot_from_gost(markers_enriched, annDB = 'GO:MF', top_res = 20, highlighted_only = T) 
plot_from_gost(markers_enriched, annDB = 'GO:MF', top_res = 20)  
#ggsave('mfplotD.png', path = filespath )

plot_from_gost(markers_enriched, annDB = 'GO:CC', top_res = 20, highlighted_only = T) 
plot_from_gost(markers_enriched, annDB = 'GO:CC', top_res = 20)

plot_from_gost(markers_enriched, annDB = 'WP', top_res = 20) 
#plot_from_gost(markers_enriched, annDB = 'KEGG', top_res = 20) 
```

a bit of exploration

```{r}
# sesequiterpene genes - related to flowering 
table(rownames(y2) == 'AT5G23960') #nope
table(rownames(y2) == 'AT5G44630')

cpm(y2$counts[which(rownames(y2) == "AT5G44630"), ])  %>% 
  data.frame(Counts = ., Genotype = y2$samples$group) %>% 
  ggboxplot(., 
          x = "Genotype", 
          y = "Counts",outlier.shape = NA,
          width = 0.5,
          title="AT5G44630", fill = "Genotype", 
          palette = group_colors[c(1,4,7,10)])# -> boxtopwt_1
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

# COMPARE WITH MAIDA 17 DAYS

```{r}
#| output: false

read_csv('DE_FD3vsWT.csv') -> results_wtfd3
#read_csv('enrichDE_FD3vsWT.csv') -> enriched_results_wtfd3

read_csv('DE_FDPvsWT.csv') -> results_wtfdp
#read_csv('enrichDE_FD3vsWT.csv') -> enriched_results_wtfdp

read_csv('DE_FD3vsWT.csv') -> results_fd3fdp
#read_csv('enrichDE_FD3vsFDP.csv') -> enriched_results_fd3fdp
```

```{r}
maida_de <- list(de_wtfd3 = (results_wtfd3 %>% unlist %>% unname),
                 de_wtfdp = (results_wtfdp %>% unlist %>% unname),
                 de_fd3fdp = (results_fd3fdp%>% unlist %>% unname))


maida_de_enrich <- list(enrich_wtfd3 = (enriched_results_wtfd3[,4]%>% 
                                          unlist %>% unname),
                        enrich_wtfdp =( enriched_results_wtfdp[,4]%>% 
                                          unlist %>% unname),
                        enrich_fd3fdp =( enriched_results_fd3fdp[,4]%>% 
                                           unlist %>% unname))
```

```{r}

# 

gom_DE <- newGOM(de_results, 
                  maida_de, 
                  dim(y2$counts)[1])

#drawHeatmap(gom.obj)

overlapDE_odds <- getMatrix(gom_DE, name = 'odds.ratio')
overlapDE_pval <- getMatrix(gom_DE, name = 'pval')

ComplexHeatmap::Heatmap(t(overlapDE_odds),
                        column_order = factor(1:3),
                        column_title = 'DEG Intersection',
                #col = rev(viridis(5, begin = 0, end = 1, option = 'magma')),
                        col = colorRamp2(c(0,20), c('#dfdfdf', 'darkgreen')),
                        cell_fun = function(j, i, x, y, width, height, fill) {
                          if(overlapDE_pval[j, i] < .05)
                             grid.text('*',
                                       x, y, 
                                       gp = gpar(fontsize = 10))
                          
                        }
                      )


```

```{r}

library(ggvenn)
# degs

results_wtareb[results_wtareb$logFC>0,] -> up_areb
results_wteel[results_wteel$logFC >0,] -> up_eel
results_wtdouble[results_wtdouble$logFC >0,] -> up_double

rbind(results_wtareb[which(results_wtareb$logFC < 0),], 
      results_wteel[which(results_wteel$logFC < 0),],
      results_wtdouble[which(results_wtdouble$logFC < 0),]) -> up_wt


ggvenn(data = list(UP_areb = up_areb[,1],
                   UP_eel = up_eel[,1],
                   UP_double = up_double[,1],
                   UP_WT = up_wt[,1]
                   ), 
      #fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
      stroke_size = 0.5, set_name_size = 4, show_percentage = F
      ) + ggtitle('Common upregulated DEGs between Damiano\'s')

ggvenn(data = list(DE_areb = results_wtareb[,1],
                   DE_eel = results_wteel[,1],
                   DE_double = results_wtdouble[,1]), 
      #fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
      stroke_size = 0.5, set_name_size = 4, show_percentage = F
      ) + ggtitle('Common DEGs (Damiano\'s mutants)')



```

```{r}
ggvenn(data = list(areb3 = results_wtareb[,1],
                   eel = results_wteel[,1],
                   fd = maida_de$de_wtfd3,
                   fdp = maida_de$de_wtfdp), 
      #fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
     stroke_size = .7,set_name_size = 4, show_percentage = F
      ) + ggtitle('DEGs similarity between mutants (Damiano and Maida)')
```

```{r}

Reduce(intersect, list(areb3 = results_wtareb[,1],
                       eel = results_wteel[,1],
                       fd = maida_de$de_wtfd3,
                       fdp = maida_de$de_wtfdp))


Reduce(intersect, list(results_wtareb[,1],
                       results_wteel[,1])) %>% data.frame

Reduce(intersect, list(results_wtareb[,1],
                       maida_de$de_wtfd3)) %>% data.frame


Reduce(intersect, list(results_wtareb[,1],
                       maida_de$de_wtfdp)) %>% data.frame


Reduce(intersect, list(results_wteel[,1],
                       maida_de$de_wtfd3)) %>% data.frame



Reduce(intersect, list(results_wteel[,1],
                       maida_de$de_wtfdp)) %>% data.frame


```

```{r}
# with DEseq

library(readxl)

read_excel("diffExp_deseqVSedger.xlsx", sheet = 1) -> deseq_wtareb
read_excel("diffExp_deseqVSedger.xlsx", sheet = 3) -> deseq_wteel
read_excel("diffExp_deseqVSedger.xlsx", sheet = 2) -> deseq_wtdbl
```

```{r}
deseq_wtareb[abs(deseq_wtareb$deseq2_log2FoldChange) >1,] # different lfc subsetting

```

```{r}

ggvenn(data = list(areb3_edger = results_wtareb[,1],
                   areb3_deseq = deseq_wtareb[,1] %>% unlist %>% unname, 
                   eel_edger = results_wteel[,1],
                   eel_deseq = deseq_wteel[,1]  %>% unlist %>% unname ), 
      #fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
     stroke_size = .7,set_name_size = 4, show_percentage = F
      ) + ggtitle('DEGs similarity between EdgeR and DESeq2')
```

```{r}

```

```{r}

```

```{r}

```

```{r}
table(results_wtareb$TAIR_ID %in% results_wteel$TAIR_ID)
results_wtareb[which(results_wtareb$TAIR_ID %in% results_wteel$TAIR_ID),-2]
results_wteel[which(results_wteel$TAIR_ID %in% results_wtdouble$TAIR_ID),-2]

cpm(y2$counts) %>% 
  data.frame %>%
  .[rownames(.) %in% c("AT3G56850", "AT2G41070"),] %>%
  stack(.) %>%
  rename(CPM = values, 
         Sample = ind) %>% 
  mutate(., Genotype = rep(group, each = 2)) %>%
  mutate(., Gene = rep(c("AT2G41070", "AT3G56850"), 11)) %>%
  ggplot(., aes(x = Gene, y = CPM, fill = Genotype)) + 
  geom_boxplot() +
  facet_grid(~Genotype) + theme_bw()



```

## 

# 
